The role of standing variation in parallel adaptation

### 2014-09-14 standing-variation-paper.tex
a bit more to discussion

The geographic area of Central and Eastern Asia with malaria is on the
order of $\sim 10$ million $km^2$. In that area there are at least $15$
common variants \citep{Howes-g6pd-variants} (see Figure \ref{fig-G6PD-map}). 
Therefore, the average width of an area occupied by an allele is $\sqrt{10^7/15} \approx  800$km. 
The coding region of G6pd is 515 codons long, 
and around 140 distinct deficiency alleles have been observed. 
Assuming a mutation rate of $\approx 10^{-8}$ per base-pair per generation, 
we can take $\mu \approx 1 \times 10^{-6}$ per generation. 
The dispersal and demographic parameters of humans in the past few thousans years is unclear,
particularly as we are concerned with the ``effective'' population density
(i.e.\ population density divided by variance in offspring number).
We therefore will use two reasonable values for the effective population density: $\rho=2$ and $0.2$ people per $km^2$,
and three values for the dispersal distance: $\sigma=10$, $50$ and $100$.
Note that obviously human migration has been shaped both by local dispersal and larger-scale expansions 
\citep[see ][for a recent discussion]{}, 
so these parameters only provide a rough view of the process.

@@

One interesting problem that we can seek to address with this work is
the extent to which the degree of pleiotropy biases 
adaptation towards the repeated use of particular subset of loci
(i.e.\ convergence at genetic level). 
While many alleles may confer the benefical phenotype, 
not all will contribute to adaptation if they have negative pleiotropic consequences.
There are at least two ways that negative pleiotropy can contribute to high rates of convergence 
when adapting if a single change is sufficient for adaptation. 
First, negative pleiotropic effects can reduce 
the overall beneficial selection coefficient of an allele in the new environment,
making them unlikely to become established and slow to spread 
(and in the worst case making them deleterious). 
This effect has been studied by a number of authors \citep{}. 
A second contribution is that alleles that have
less negative pleiotropy are more likely to be present as standing variation 
before the environmental shift.
This effect seems less well studied, 
perhaps because many modeling approaches use the strong selection, weak mutation limit 
and so ignore standing variation. \gc{Is this fair?}
\plr{well, even in the strong-selection weak-mutation limit you'll have standing variation}

@@

To illustrate some of the properties of this model, 
let's imagine the somewhat extreme scenerio that there is a single base-pair 
at which a possible mutation is relatively free of negative pleiotropy (call this class $1$), 
and a larger mutational target where changes 
have more serious pleiotropic consequences in the ancestral environment (class $2$). 
We set $s_{d,1} \leq s_{d,2}$ and $\mu_1=1 \times 10^{-8}$, 
assume that both classes share a beneficial selection coefficient of $s_b=0.05$,
and think of the second class mutational target being ten, a hundred, or one thousand base pairs. 
\plr{What is $s_{d,2}$ in figure 7?}
We show the expected proportion of space covered by
the rarer, class $1$ mutations in Figure \ref{fig-pleiotropy_calc}.
Intuitively, the contribution of the rarer mutation decreases as the
mutational target of the other class becomes larger, and as the
difference in the negative pleiotropic consequences of the two classes of alleles decreases. 
The case with standing variation only is the best case scenario 
for the rarer mutation as we force them to have the same advantage in the new environment, 
so its rate of introduction is necessarily lower. 
However, the standing variation only case does seem to provide a reasonable rule of thumb. 
Especially for parameter combinations, 
higher population densities and high dispersal distances, 
that increase the contribution of standing variation (and similarly for high $s_b$).

@@

\paragraph{Are species held together by wide-spread selective sweeps?}
Our results touch on an old debate on the evolutionary coherence of species. 
Mayr and many others have argued that species are coherent
evolutionary units because they are united by shared gene flow \citep[pages 521–522 ][]{Mayr:SpeciesEvol}. 
However, this argument has been questioned by a number of authors 
based on relatively high levels of differentiation, and low rates of dispersal, in many species \citep{EhrlichRaven:69,Levin:79}.
\citet{Rieseberg2001} and \citet{MorjanRieseberg:04} argued that 
even if gene flow is not be high enough to prevent neutral differentiation, 
species should be viewed as being united 
if levels of gene flow were still high enough for selected alleles to spread across the entire species \citep{Ellstrand2014s}.
\plr{not clear which of three papers in the last sentence are being cited for what?}
At present, large scale genotyping and sequencing projects, 
along with more sophisticated methods, 
are highlighting ever more signals of gene flow between populations and species \citep{}.
However, our work on geographic convergent adaptation
\citep[see also][]{ralph2010parallel,RalphCoop:14} 
concludes that this gene flow will lead to the rapid spread of
selected alleles partially into question. 
Our results suggest that species should often adapt to
wide-spread selection pressures through convergent evolution rather
than waiting for a single allele to migrate across the range.
% over length-scales determined by our characteristic length.

@@

>>>>

The geographic area of Central and Eastern Asia with malaria is 
on the order of ten million square kilomters.
In that area there are at least $15$ common variants \citep{Howes-g6pd-variants} (see Figure \ref{fig-G6PD-map}). 
Therefore, the average width of an area occupied by an allele is $\sqrt{10^7/15} \approx  800$km. 
The coding region of G6pd is 515 codons long, 
and around 140 distinct deficiency alleles have been observed. 
Assuming a mutation rate of $\approx 10^{-8}$ per base-pair per generation, 
we can take $\mu \approx 1 \times 10^{-6}$ per generation. 
The dispersal and demographic parameters of humans in the past few thousans years is unclear,
particularly as we are concerned with the ``effective'' population density
(i.e.\ population density divided by variance in offspring number).
We therefore will use two reasonable values for the effective population density: $\rho=2$ and $0.2$ people per $km^2$,
and three values for the dispersal distance: $\sigma=10$, $50$ and $100$.
Note that obviously human migration has been shaped both by local dispersal and larger-scale expansions 
\citep[see ][for a recent discussion]{}, 
so these parameters only provide a rough view of the process.

@@

One interesting problem that we can seek to address with this work is
the extent to which the degree of pleiotropy biases 
adaptation towards the repeated use of particular subset of loci
(i.e.\ convergence at genetic level). 
While many alleles may confer the benefical phenotype, 
not all will contribute to adaptation if they have negative pleiotropic consequences.
There are at least two ways that negative pleiotropy can contribute to high rates of convergence 
when adapting if a single change is sufficient for adaptation. 
First, negative pleiotropic effects can reduce 
the overall beneficial selection coefficient of an allele in the new environment,
making them unlikely to become established and slow to spread 
(and in the worst case making them deleterious). 
This effect has been studied by a number of authors \citep{}. 
A second contribution is that alleles that have
less negative pleiotropy are more likely to be present as standing variation 
before the environmental shift.
This effect seems less well studied, 
perhaps because many modeling approaches use the strong selection, weak mutation limit 
and so ignore standing variation. 
\gc{Is this fair?}
\plr{Not sure what papers you are thinking of.}

@@

To illustrate some of the properties of this model, 
let's imagine the somewhat extreme scenario that there is a single base-pair 
at which a possible mutation is relatively free of negative pleiotropy (call this class $1$), 
and a larger mutational target where changes 
have more serious pleiotropic consequences in the ancestral environment (class $2$). 
We set $s_{d,1} \leq s_{d,2}$ and $\mu_1=1 \times 10^{-8}$, 
assume that both classes share a beneficial selection coefficient of $s_b=0.05$,
and think of the second class mutational target being ten, a hundred, or one thousand base pairs. 
\plr{What is $s_{d,2}$ in figure 7?}
We show the expected proportion of space covered by
the rarer, class $1$ mutations in Figure \ref{fig-pleiotropy_calc}.
Intuitively, the contribution of the rarer mutation decreases as the
mutational target of the other class becomes larger, and as the
difference in the negative pleiotropic consequences of the two classes of alleles decreases. 
The case with standing variation only is the best case scenario 
for the rarer mutation as we force them to have the same advantage in the new environment, 
so its rate of introduction is necessarily lower. 
However, the standing variation only case does seem to provide a reasonable rule of thumb. 
Especially for parameter combinations, 
higher population densities and high dispersal distances, 
that increase the contribution of standing variation (and similarly for high $s_b$).

@@

\paragraph{Are species held together by widespread selective sweeps?}
Our results touch on an old debate on the evolutionary coherence of species. 
Mayr and many others have argued that species are coherent
evolutionary units because they are united by shared gene flow \citep[pages 521--522 ][]{Mayr:SpeciesEvol}. 
However, this argument has been questioned by a number of authors 
based on relatively high levels of differentiation, and low rates of dispersal, in many species \citep{EhrlichRaven:69,Levin:79}.
\citet{Rieseberg2001} and \citet{MorjanRieseberg:04} argued that 
even if gene flow is not be high enough to prevent neutral differentiation, 
species should be viewed as being united 
if levels of gene flow were still high enough for selected alleles to spread across the entire species \citep{Ellstrand2014s}.
\plr{not clear which of three papers in the last sentence are being cited for what?}
At present, large scale genotyping and sequencing projects, 
along with more sophisticated methods, 
are highlighting ever more signals of gene flow between populations and species \citep{}.
However, our work on geographic convergent adaptation
\citep[see also][]{ralph2010parallel,RalphCoop:14} 
concludes that this gene flow will lead to the rapid spread of
selected alleles partially into question. 
Our results suggest that species should often adapt to
wide-spread selection pressures through convergent evolution rather
than waiting for a single allele to migrate across the range.
% over length-scales determined by our characteristic length.

@@

\plr{added:}
Even if selective sweeps only bring alleles to fixation locally,
they are still potentially a stronger homogenizing force
than neutral mixing through migration.
We can see this by comparing mean coalescent times.
Under neutral mixing, the mean number of generations back to the most recent common ancestor
is on order of the total population size.
This quantity has not been worked out for a model with simultaneous local sweeps,
but is close to the ``spatial $\Lambda$-Fleming--Viot'' models of \citet{barton2013modelling},
in which local sweeps occur independently across the range.
In these models, the mean number of generations back to the most recent common ancestor for distant samples
scales with XXX,
which can be substantially shorter.
\plr{Maybe I'm forgetting something in their papers -- did they do this? --
but I think it would be a little bit of work to figure out what parameters to put into their models.}
Intuitively, we can obtain this as follows:
suppose that coalescence only occurs during local sweeps,
which occur at rate $\lambda$
on a spatial scale such that there are on average $M$ independent origins of the sweep during each bout of adaptation.
The motion of lineages backwards is not dissimilar to a stepping-stone model with $M$ sites
that coalesce when they are in the same deme
with probability dependent on the recombination distance to the sweeping, selected site.
The mean hitting time of two independent random walks on a graph with $M$ nodes is of order $M$;
to get the mean number of generations until coalescence, 
we must divide this by the probability of remaining linked to the selected site as it sweeps
and by the frequency of selected sites.
Since $M$ scales proportionally to population density $\rho$,
coalescent times do as well,
but possibly with a much smaller coefficient
(due effectively to an increase in reproductive variance).
